Rapid technological progress in white light-emitting diodes and its sources in innovation and technology spillovers

### 2023-08-17 2_SSL_NE/article/article_main.tex
added edits

@@

\cref{fgr:subeff} shows the overall magnitude of contributions of identified LED innovations and technology spillovers to improvements in LED efficiency over time across different sub-efficiencies. The list of corresponding technology spillovers is provided in \cref{tab:spillovers} in \cref{sec:discussion}. Through the index decomposition analysis described in \cref{subsec:performance_metrics}, we find that out of the overall LED efficiency increase of 32.9\% from 5.8\% to 38.8\% between 2003 and 2020, at least 2.8\% can be attributed specifically to innovations that involved technology spillovers identified in this study, corresponding to 8.5\% of the total LED efficiency improvements between 2003 and 2020.

@@

\begin{figure}[h!]
 \centering
 \includegraphics[width=15.5cm]{figures/waterfall_performance_2003.pdf}
 \includegraphics[width=15.5cm]{figures/waterfall_performance_2010.pdf}
 \includegraphics[width=15.5cm]{figures/waterfall_performance_2016.pdf}
 \includegraphics[width=15.5cm]{figures/waterfall_performance_2020.pdf}
 \captionsetup{font=footnotesize}
 \caption{\textbf{Waterfall diagrams of the loss channels in a generic mid/high-power LED package for 2003, 2010, 2016 and 2020} (top to bottom panels). Losses are normalized to 1 Watt of electric power input (yellow bar on the left). Sub-efficiencies corresponding to each loss channel are listed below each column and described in Supplementary Note 2.  Numbers for each loss channel indicate energy losses both in relative terms of input power (in percent) at the point of the channel and absolute values (in Watts). Percentages for loss channels labeled by red, green and blue font indicate losses of corresponding remaining red/green/blue light energy. The following LED architectures and down-conversion phosphor spectra used in calculations in each considered year are shown for reference: 2003 – Flip-chip with YGAG phosphor; 2010 – Flip-chip with 258 phosphor; 2016 – Flip-chip with PFS phosphor; 2020 – Chip-scale package flip-chip with SALON phosphor. Details for each architecture are provided in \cref{fgr:chip_architecture_overview}.  Details on the phosphors for light down conversion are provided in \cref{tab:phosphors} and Supplementary Note 4. Abbreviations: Scat. = Scattering; Nonrad. = Nonradiative. Source: own elaboration based on data on sub-efficiencies presented in Supplementary Figure 9 and spectral data in Supplementary Figure 10,  adapted from publications on the respective phosphors: YGAG (2003)\cite{Mueller2002}, 258 (2010)\cite{MuellerMach2005}, PFS (2016)\cite{Murphy2015}, SALON (2019)\cite{Hoerder2019}.}
 \label{fgr:waterfall}
\end{figure}

\begin{figure}[h!]
 \centering
 \includegraphics[width=\textwidth]{figures/breakthroughs_efficiency.pdf}
 \caption{\textbf{Contribution of innovations and technology spillovers to the historical progress in sub-efficiencies of phosphor-converted warm white LEDs.} LEDs with test currents of at least 350mA are considered. The following sub-efficiencies are represented: $\eta_{V_f}$ - forward voltage efficiency; $\eta_IQ$ - internal quantum efficiency; $\eta_{droop}$ - efficiency droop; $\eta_{LE}$ - light extraction efficiency; $\eta_{C_R}$ - conversion efficiency for red phosphors; $\eta_{C_{Y/G}}$ - conversion efficiency for yellow/green phosphors; $\eta_S$ - spectral efficiency. The overall LED lamp efficiency $\eta_L$ is displayed in the rightmost column. Vertical bars represent cumulative contributions to efficiency improvements from LED technology innovations, with purple bars indicating innovations driven by technology spillovers (annotated and listed in an inset table) and grey bars indicating all other improvements identified in this study. Horizontal coloured lines indicate sub-efficiency levels of best performing LEDs for the four years used in \cref{fgr:waterfall}: 2003, 2010, 2016, and 2020. For additional historical context, data for 1997 is included whenever possible. "N/A" denotes sub-efficiencies where 1997 data could not be calculated for the following reasons: $\eta_{V_f}, \eta_{droop}$ depend on the device current, which was below 350mA in 1997, making a comparison with contemporary devices difficult. $\eta_{C_R}, \eta_{C_{Y/G}}$ and $\eta_S$ are relevant only to warm white spectrum LEDs, which were not available in 1997. Note: ITO current spreading layer affects different sub-efficiencies in different chip architectures, e.g., in modern flip-chip architectures $\eta_{LE}$ no longer depends on ITO, see \cref{fgr:chip_architecture_overview}. Physical limits on sub-efficiencies are indicated by black horizontal lines. Note that the physical limit of $\eta_{V_f}$ is above 100\% due to quantum effects \cite{david2016electrical}. Source: own elaboration based on data represented in \cref{fgr:waterfall}, \cref{tab:spillovers} and Supplementary Note 4.}
 \label{fgr:breakthroughs_efficiency}
\end{figure}

@@

\begin{table}[h!]
    \footnotesize
    \centering
    \caption{\textbf{LED innovations related to improvements in consumer experience metrics.}}
    \begin{NiceTabularX}{\textwidth}{|l|l|l|X|X|l|}
    \hline
        \textbf{Year} & \textbf{Desig.} & \textbf{Chemical formula} & \textbf{Description} & \textbf{Significance} & \textbf{Ref.} \\ \hline
        1996 & YAG & $Y_3 Al_5 O_{12}:Ce$ & Yttrium aluminium garnet (YAG) phosphor activated with cerium & First LED phosphor, enabled white LEDs & SI 3.2.2.1 \\ \hline
        1996 & YGAG & $(Y_{1-x} Gd_x)_3 Al_5 O_{12}:Ce$ & Gadolinium-doped YAG phosphor & First red-shifted phosphor, enabled warm white LEDs & SI 3.2.2.1 \\ \hline
        2002 & 258 & $(Ba,Sr)_2 Si_5 N_8:Eu^{2+}$ & Europium-doped nitridosilicate phosphor & First red LED phosphor & SI 3.2.2.2 \\ \hline
        2003 & QD & N/A & Quantum dot-based phosphor & First use of QD for LED light down conversion & SI 3.2.2.4 \\ \hline
        2005 & PFS & $K_2 SiF_6: Mn^{4+}$ & Manganese-activated potassium fluorosilicate (PFS) phosphor & First ultra-narrow-band red LED phosphor & SI 3.2.2.3 \\ \hline
        2013 & SLA & $Sr[Li Al]_3 N_4 ]:Eu^{2+}$ & Europium-doped cuboidal nitridolithoaluminate phosphor & Improved narrow-band red phosphor & SI 3.2.2.2 \\ \hline
        2016 & SALON & $Sr[Li_2 Al_2 O_2 N_2]:Eu^{2+}$ & Europium-doped oxonitride phosphor & High-performance ultra-narrow-band red phosphor & SI 3.2.2.2 \\ \hline
    \end{NiceTabularX}
    \caption*{Note: \textit{Year} column represents the earliest reported application of innovation in white LEDs. These differ from the years shown in \cref{fgr:consumer_experience} which correspond to the earliest publication of spectral data for LED products that relied on those innovations. Detailed descriptions of the history of each innovation and relevant spectral data are provided in Supplementary Note 4. Desig. = Designation, Ref. - Reference.}
    \label{tab:phosphors}
\end{table}

\begin{figure}[h!]
 \centering
 \includegraphics[width=\textwidth]{figures/breakthroughs_consumer-experience.pdf}
 \caption{\textbf{Historical improvements in consumer experience metrics of phosphor-converted white LEDs.} Data points show the color temperature (CT) and colour rendering index (CRI) performance of the earliest identified representative white LED products with published spectral data that used phosphors listed in \cref{tab:phosphors}, each indicated by the phosphor label and publication year. Corresponding spectral data used to calculate CT values is provided in Supplementary Note 4. Three data points representing compact fluorescent lamps (CFLs)\cite{cie_reference}, shown as black triangles, are provided for comparison. Horizontal lines represent typical CTs of “traditional” light sources, shown for reference. The desirable range of colour temperatures for home illumination, indicated by a vertical red arrow, lies between two horizontal orange lines representing typical incandescent light and warm daylight colour temperatures. Horizontal red arrow indicates desirable higher values of colour rendering index (CRI). LED products based on the following phosphor innovations from the following LED manufacturers are represented : YAG - Nichia, 1998 \cite{bando1998development}; YGAG - Lumileds, 2002 \cite{Mueller2002}; 258 - Lumileds, 2005 \cite{MuellerMach2005}; SLA - Lumileds, 2014 \cite{Pust2014}; PFS - GE, 2015 \cite{Murphy2015}; QD - Lumileds, 2017 \cite{lumileds2016qd}, Osram, 2019 \cite{osram2019qd}; SALON - Osram, 2019 \cite{Hoerder2019}.}
 \label{fgr:consumer_experience}
\end{figure}

@@

\cref{fgr:costmodel} shows main results of our manufacturing cost modelling for low-to-mid-power classic chip GaN phosphor-converted white LED packages. We find that the manufacturing cost of a single such LED decreased from 1.11\$ (in 2020 USD) in 2003 to 0.11\$ (2020 USD) in 2012 and 0.05\$ in 2020, a 95.5\% overall decrease.

@@

\begin{figure*}[h!]
 \centering
 \includegraphics[width=17.5cm]{figures/costmodel_results_years.pdf}
 \caption{\textbf{Historical changes in LED manufacturing cost structure.} Presented cost calculations are an outcome of our cost model (see \cref{sec:methods} and \cref{costmodel_details}) for a single low-to-mid power GaN-based, classic chip, phosphor-converted white LED package, assuming an ideal factory with state-of-the-art equipment at a U.S. location. Panels A-C: Waterfall diagrams of LED manufacturing cost structure split by manufacturing process steps for years 2003, 2012 and 2020. Process steps on the horizontal axis are sequenced from left to right in the same order as in the modelled LED manufacturing process. Panel D: Cumulative manufacturing yield after each process step for years 2003, 2012 and 2020. For a graphic representation of the manufacturing process, see the diagrams in \cref{subsec:manufacturing_steps}. Abbreviations: Litho – Lithographic Process, Insp. – Inspection, Depn. – Deposition, CMP - Chemical-Mechanical Planarization.}
 \label{fgr:costmodel}
\end{figure*}

@@

Our LED manufacturing cost model shows a 95.5\% decrease in the cost of producing low-to-mid power GaN-based white classic-chip LEDs from 1.11\$ to 0.05\$ (in 2020 USD) between 2003 and 2020. In contrast with LED performance improvements, where progress was driven mostly by innovations in LED technology, the dramatic decline in LED manufacturing cost has been a product of higher yields across manufacturing steps and economies of scale resulting from increases in the sapphire wafer size. Surprisingly, LED innovations and associated technology spillovers are not found to have a substantial impact on LED manufacturing cost reductions over time. There are two reasons for this:

First, the \textit{functional unit} of our study is the LED chip itself, not the luminaire containing multiple chips. As efficiency and overall LED brightness has increased, a smaller number of chips could be used in luminaires to obtain the same level of overall flux. However, since we are investigating improvements at the level of single chips, we do not capture cost effects downstream of chip manufacturing. Nonetheless, we were previously able to show that overall brightness is no longer a target metric in chip development \cite{weinold2021compound}.

Second, we find that an increase in wafer size and higher overall yield together are responsible for a 98\% reduction of manufacturing cost per LED die between between 2003 and 2020. Both were enabled by advances in manufacturing equipment (e.g., for epitaxy and chip dicing) and incremental manufacturing process improvements from learning-by-doing, which was confirmed by multiple interviews with experts from manufacturing. A plausible explanation for this finding is that the main focus of R&D activities in white LED technology in the considered period was likely on LED performance improvements rather than cost reductions.

@@

There are several important avenues of future research that are opened up by our analysis. First, future work could expand the cost model by collecting and including data for a broader set of chip architectures. Second, a deeper dive into the role of learning-by-doing is needed both in the cost and performance analysis. Third, building on the work on the physical limits in LED sub-efficiencies, future efforts could focus on identifying priority areas for further performance improvements in LEDs and SSL in general. Fourth, better understanding is needed about the role of demand-side factors, such as policies stimulating market demand for LED lighting (e.g., through incandescent light bulb bans or subsidies for LED adoption), not only in the rapid diffusion of LED-based lighting technologies, which is reasonably well understood by now \cite{Mills2014, Kamat2020, weinold2021quantifying, stegmaier2021incandescent, grubb2021new}, but also in facilitating learning-by-doing, innovation and technology spillovers that have contributed to the technological progress in white LEDs. Finally, the lessons learned about the drivers of innovation, technology spillovers, cost reductions and performance improvements in LEDs may have broader implications both for other demand-side or energy-efficiency technologies, and for low-carbon energy technologies with similar characteristics such as the granularity of technology, its modularity, and complexity \cite{malhotra2020accelerating, Wilson2012}. By comparing these factors at a granular level across different low-carbon and energy efficiency technologies, we can identify and generalize key cross-technology patterns or differences that would help us formulate recommendations for industry and policymakers aimed at accelerating further clean energy innovation for climate change mitigation.

\section{Methods}

@@

To overcome the limitations of existing literature and patent analysis, in this study we rely on a multi-method approach to data collection and analysis, augmenting information obtained from a systematic review of the primary scientific literature, device datasheets, relevant patents, and industry publications with information gained from semi-structured interviews with experts from academia and industry, bottom-up manufacturing cost modelling, and own calculations of LED device sub-efficiencies. We then synthesize information obtained with multiple methods to track the historical progress in white LED technology over time across the three groups of metrics (see Supplementary Note 2) and identify its sources in innovation and technology spillovers. We briefly describe our approach to each method below and provide further methodological details in Supplementary Note 3.

>>>>

% number all lines as pre journal submission guidelines
\usepackage{lineno}
\linenumbers

@@

\cref{fgr:breakthroughs_efficiency} shows the overall magnitude of contributions of identified LED innovations and technology spillovers to improvements in LED efficiency over time across different sub-efficiencies. The list of corresponding technology spillovers is provided in \cref{tab:spillovers} in \cref{sec:discussion}. Through the index decomposition analysis described in \cref{subsec:performance_metrics}, we find that out of the overall LED efficiency increase of 32.9\% from 5.8\% to 38.8\% between 2003 and 2020, at least 2.8\% can be attributed specifically to innovations that involved technology spillovers identified in this study, corresponding to 8.5\% of the total LED efficiency improvements between 2003 and 2020.

@@

\begin{figure}[H]
 \centering
 \includegraphics[width=15.5cm]{figures/waterfall_performance_2003.pdf}
 \includegraphics[width=15.5cm]{figures/waterfall_performance_2010.pdf}
 \includegraphics[width=15.5cm]{figures/waterfall_performance_2016.pdf}
 \includegraphics[width=15.5cm]{figures/waterfall_performance_2020.pdf}
 \captionsetup{font=footnotesize}
 \caption{\textbf{Waterfall diagrams of the loss channels in a generic mid/high-power LED package for 2003, 2010, 2016 and 2020} (top to bottom panels). Losses are normalized to 1 Watt of electric power input (yellow bar on the left). Sub-efficiencies corresponding to each loss channel are listed below each column and described in Supplementary Note 2.  Numbers for each loss channel indicate energy losses both in relative terms of input power (in percent) at the point of the channel and absolute values (in Watts). Percentages for loss channels labeled by red, green and blue font indicate losses of corresponding remaining red/green/blue light energy. The following LED architectures and down-conversion phosphor spectra used in calculations in each considered year are shown for reference: 2003 – Flip-chip with YGAG phosphor; 2010 – Flip-chip with 258 phosphor; 2016 – Flip-chip with PFS phosphor; 2020 – Chip-scale package flip-chip with SALON phosphor. Details for each architecture are provided in \cref{fgr:chip_architecture_overview}.  Details on the phosphors for light down conversion are provided in \cref{tab:phosphors} and Supplementary Note 4. Abbreviations: Scat. = Scattering; Nonrad. = Nonradiative. Source: own elaboration based on data on sub-efficiencies presented in Supplementary Figure 9 and spectral data in Supplementary Figure 10,  adapted from publications on the respective phosphors: YGAG (2003)\cite{Mueller2002}, 258 (2010)\cite{MuellerMach2005}, PFS (2016)\cite{Murphy2015}, SALON (2019)\cite{Hoerder2019}.}
 \label{fgr:waterfall}
\end{figure}

\begin{figure}[H]
 \centering
 \includegraphics[width=\textwidth]{figures/breakthroughs_efficiency.pdf}
 \caption{\textbf{Contribution of innovations and technology spillovers to the historical progress in sub-efficiencies of phosphor-converted warm white LEDs.} LEDs with test currents of at least 350mA are considered. The following sub-efficiencies are represented: $\eta_{V_f}$ - forward voltage efficiency; $\eta_IQ$ - internal quantum efficiency; $\eta_{droop}$ - efficiency droop; $\eta_{LE}$ - light extraction efficiency; $\eta_{C_R}$ - conversion efficiency for red phosphors; $\eta_{C_{Y/G}}$ - conversion efficiency for yellow/green phosphors; $\eta_S$ - spectral efficiency. The overall LED lamp efficiency $\eta_L$ is displayed in the rightmost column. Vertical bars represent cumulative contributions to efficiency improvements from LED technology innovations, with purple bars indicating innovations driven by technology spillovers (annotated and listed in an inset table) and grey bars indicating all other improvements identified in this study. Horizontal coloured lines indicate sub-efficiency levels of best performing LEDs for the four years used in \cref{fgr:waterfall}: 2003, 2010, 2016, and 2020. For additional historical context, data for 1997 is included whenever possible. "N/A" denotes sub-efficiencies where 1997 data could not be calculated for the following reasons: $\eta_{V_f}, \eta_{droop}$ depend on the device current, which was below 350mA in 1997, making a comparison with contemporary devices difficult. $\eta_{C_R}, \eta_{C_{Y/G}}$ and $\eta_S$ are relevant only to warm white spectrum LEDs, which were not available in 1997. Note: ITO current spreading layer affects different sub-efficiencies in different chip architectures, e.g., in modern flip-chip architectures $\eta_{LE}$ no longer depends on ITO, see \cref{fgr:chip_architecture_overview}. Physical limits on sub-efficiencies are indicated by black horizontal lines. Note that the physical limit of $\eta_{V_f}$ is above 100\% due to quantum effects \cite{david2016electrical}. Source: own elaboration based on data represented in \cref{fgr:waterfall}, \cref{tab:spillovers} and Supplementary Note 4.}
 \label{fgr:breakthroughs_efficiency}
\end{figure}

@@

\begin{table}[H]
    \footnotesize
    \centering
    \caption{\textbf{LED innovations related to improvements in consumer experience metrics.}}
    \begin{NiceTabularX}{\textwidth}{|l|l|l|X|X|l|}
    \hline
        \textbf{Year} & \textbf{Desig.} & \textbf{Chemical formula} & \textbf{Description} & \textbf{Significance} & \textbf{Ref.} \\ \hline
        1996 & YAG & $Y_3 Al_5 O_{12}:Ce$ & Yttrium aluminium garnet (YAG) phosphor activated with cerium & First LED phosphor, enabled white LEDs & SI 3.2.2.1 \\ \hline
        1996 & YGAG & $(Y_{1-x} Gd_x)_3 Al_5 O_{12}:Ce$ & Gadolinium-doped YAG phosphor & First red-shifted phosphor, enabled warm white LEDs & SI 3.2.2.1 \\ \hline
        2002 & 258 & $(Ba,Sr)_2 Si_5 N_8:Eu^{2+}$ & Europium-doped nitridosilicate phosphor & First red LED phosphor & SI 3.2.2.2 \\ \hline
        2003 & QD & N/A & Quantum dot-based phosphor & First use of QD for LED light down conversion & SI 3.2.2.4 \\ \hline
        2005 & PFS & $K_2 SiF_6: Mn^{4+}$ & Manganese-activated potassium fluorosilicate (PFS) phosphor & First ultra-narrow-band red LED phosphor & SI 3.2.2.3 \\ \hline
        2013 & SLA & $Sr[Li Al]_3 N_4 ]:Eu^{2+}$ & Europium-doped cuboidal nitridolithoaluminate phosphor & Improved narrow-band red phosphor & SI 3.2.2.2 \\ \hline
        2016 & SALON & $Sr[Li_2 Al_2 O_2 N_2]:Eu^{2+}$ & Europium-doped oxonitride phosphor & High-performance ultra-narrow-band red phosphor & SI 3.2.2.2 \\ \hline
    \end{NiceTabularX}
    \caption*{Note: \textit{Year} column represents the earliest reported application of innovation in white LEDs. These differ from the years shown in \cref{fgr:consumer_experience} which correspond to the earliest publication of spectral data for LED products that relied on those innovations. Detailed descriptions of the history of each innovation and relevant spectral data are provided in Supplementary Note 4. Desig. = Designation, Ref. - Reference.}
    \label{tab:phosphors}
\end{table}

\begin{figure}[H]
 \centering
 \includegraphics[width=\textwidth]{figures/breakthroughs_consumer-experience.pdf}
 \caption{\textbf{Historical improvements in consumer experience metrics of phosphor-converted white LEDs.} Data points show the color temperature (CT) and colour rendering index (CRI) performance of the earliest identified representative white LED products with published spectral data that used phosphors listed in \cref{tab:phosphors}, each indicated by the phosphor label and publication year. Corresponding spectral data used to calculate CT values is provided in Supplementary Note 4. Three data points representing compact fluorescent lamps (CFLs)\cite{cie_reference}, shown as black triangles, are provided for comparison. Horizontal lines represent typical CTs of “traditional” light sources, shown for reference. The desirable range of colour temperatures for home illumination, indicated by a vertical red arrow, lies between two horizontal orange lines representing typical incandescent light and warm daylight colour temperatures. Horizontal red arrow indicates desirable higher values of colour rendering index (CRI). LED products based on the following phosphor innovations from the following LED manufacturers are represented : YAG - Nichia, 1998 \cite{bando1998development}; YGAG - Lumileds, 2002 \cite{Mueller2002}; 258 - Lumileds, 2005 \cite{MuellerMach2005}; SLA - Lumileds, 2014 \cite{Pust2014}; PFS - GE, 2015 \cite{Murphy2015}; QD - Lumileds, 2017 \cite{lumileds2016qd}, Osram, 2019 \cite{osram2019qd}; SALON - Osram, 2019 \cite{Hoerder2019}.}
 \label{fgr:consumer_experience}
\end{figure}

@@

\cref{fgr:costmodel} shows main results of our manufacturing cost modelling for low-to-mid-power classic chip GaN phosphor-converted white LED packages. We find that the manufacturing cost of a single such LED decreased from 1.11\$ (in 2020 USD) in 2003 to 0.11\$ (2020 USD) in 2012 and 0.05\$ in 2020, a 95.5\% overall decrease. Our model shows that for the wafer-level manufacturing steps illustrated in Supplementary Figures 2-6, yield improvements and an increase in  wafer size together are responsible for a 98\% reduction of manufacturing cost per LED chip between between 2003 and 2020.

@@

\begin{figure}[ht!]
\centering
\includegraphics[width=17.5cm]{figures/costmodel_results_years.pdf}
\caption{\textbf{Historical changes in LED manufacturing cost structure.} Presented cost calculations are an outcome of our cost model (see \cref{sec:methods} and (see Methods and Supplementary Note 3) for a single low-to-mid power GaN-based, classic chip, phosphor-converted white LED package, assuming an ideal factory with state-of-the-art equipment at a U.S. location. Panels A-C: Waterfall diagrams of LED manufacturing cost structure split by manufacturing process steps for years 2003, 2012 and 2020. Process steps on the horizontal axis are sequenced from left to right in the same order as in the modelled LED manufacturing process. Panel D: Cumulative manufacturing yield after each process step for years 2003, 2012 and 2020. For a graphic representation of the manufacturing process, see the diagrams in Supplementary Figures 2-6. Abbreviations: Litho – Lithographic Process, Insp. – Inspection, Depn. – Deposition, CMP - Chemical-Mechanical Planarization.}
\label{fgr:costmodel}
\end{figure}

@@

Our LED manufacturing cost model shows a 95.5\% decrease in the cost of producing low-to-mid power GaN-based white classic-chip LEDs from 1.11\$ to 0.05\$ (in 2020 USD) between 2003 and 2020. In contrast with LED performance improvements, where progress was driven mostly by innovations in LED technology, the dramatic decline in LED manufacturing cost has been a product of higher yields across manufacturing steps and economies of scale resulting from increases in the sapphire wafer size.

Surprisingly, LED innovations and associated technology spillovers are not found to have a substantial impact on LED manufacturing cost reductions over time. There are several explanations for this finding. First, the functional unit of our study is the LED chip itself, not the luminaire containing multiple chips. As LED efficiency and overall brightness have increased, a smaller number of chips can be used in luminaires to obtain the same level of the overall light flux. However, since we are investigating improvements at the level of single chips, we do not capture the cost effects downstream of chip manufacturing. We also previously showed that overall brightness is no longer a target metric in chip development \cite{weinold2021compound}.

Second, we find that an increase in wafer size and higher overall yield, taken together, are jointly responsible for a 98\% reduction of manufacturing cost per LED chip between between 2003 and 2020. Both were enabled by advances in manufacturing equipment (e.g., for epitaxy and chip dicing) and incremental manufacturing process improvements from learning-by-doing, rather than particular LED innovations, which was confirmed by interviews with experts in LED manufacturing.

@@

Finally, the lessons learned about the drivers of innovation, technology spillovers, cost reductions and performance improvements in LEDs may have broader implications both for other demand-side or energy-efficiency technologies, and for low-carbon energy technologies with similar characteristics such as the granularity of technology, its modularity, and complexity \cite{malhotra2020accelerating, Wilson2012}. By comparing these factors at a granular level across different low-carbon and energy efficiency technologies, we can identify and generalize key cross-technology patterns or differences that would help us formulate recommendations for industry and policymakers aimed at accelerating further clean energy innovation for climate change mitigation.

There are several important avenues of future research that are opened up by our analysis. First, future work could expand the cost model by collecting and including data for a broader set of chip architectures. Second, a deeper dive into the role of learning-by-doing is needed both in the cost and performance analysis. Third, building on the work on the physical limits in LED sub-efficiencies, future efforts could focus on identifying priority areas for further performance improvements in LEDs and SSL in general. Fourth, better understanding is needed about the role of demand-side factors, such as policies stimulating market demand for LED lighting (e.g., through incandescent light bulb bans or subsidies for LED adoption), not only in the rapid diffusion of LED-based lighting technologies, which is reasonably well understood by now \cite{Mills2014, Kamat2020, weinold2021quantifying, stegmaier2021incandescent, grubb2021new}, but also in facilitating learning-by-doing, innovation and technology spillovers that have contributed to the technological progress in white LEDs.

\section{Methods}
\label{sec:methods}

@@

To overcome the limitations of existing literature and patent analysis, in this study we rely on a multi-method approach to data collection and analysis, augmenting information obtained from a systematic review of the primary scientific literature \cite{Haddaway_2014}, device datasheets, relevant patents, and industry publications with information gained from semi-structured interviews with experts from academia and industry, bottom-up manufacturing cost modelling, and own calculations of LED device sub-efficiencies. We then synthesize information obtained with multiple methods to track the historical progress in white LED technology over time across the three groups of metrics (see Supplementary Note 2) and identify its sources in innovation and technology spillovers. We briefly describe our approach to each method below and provide further methodological details in Supplementary Note 3.
